\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-07-30 papers/methods/paper.tex
Adding a bit more clarification from dans emails

Like most of the BHM methods introduced previously, we work from the summary statistics, where each observed supernova has a flux measurement $\hat{m}_B$ (which is analogous to apparent magnitude), stretch $\hat{x}_1$ and colour $\hat{c}$, with uncertainty $\cov$ on those values. Additionally, each supernova has an observed redshift $\hat{z}$ and a host galaxy mass associated with it, $\hat{m}$, where the mass measurement takes the form of a probability of being above $10^{10}$ solar masses. We will also have a probability of each supernovae being a Type Ia, $\hat{p}$. Our set of observables is therefore given as $\lbrace \hat{m}_B, \hat{x}_1, \hat{c}, \hat{z}, \hat{p}, \hat{m}, \cov \rbrace$, as shown in the probabilistic graphical model (PGM) in Figure \ref{fig:pgm}.

@@

@@

where $\mathcal{C}$ is a covariance matrix reported from light curve fitting. Unfortunately, this assumption of Gaussianity is incorrect. The intrinsic supernovae population is commonly described in terms of coherent magnitude dispersion, wherein some supernovae are uniformly brighter or dimmer than the population mean, and in chromatic dispersion, whereby the spectral energy density of the supernovae varies non-uniformly as a function of wavelength. The presence of chromatic smearing in the intrinsic supernova population \citep{Guy2010, Chotard2011} means the summary statistics reported from light curve fitting are under-reported, not Gaussian and equation \eqref{eq:pop} is violated. We address this problem with a similar method to previous analytic likelihoods: including extra dispersion in the reported covariance matrix $\mathcal{C}$. Whilst this doesn't solve the issue of non-Gaussianity, it does address the under-reporting of uncertainty, which is the primary source of bias.

We model the extra dispersion only in colour, and do so by adding extra independent uncertainty on the colour observation. Fully covariant extra dispersion on $\lbrace m_B, x_1, c \rbrace$ was also tested, but it showed negligible improvement in recovering unbiased cosmology over just colour dispersion, and was far more computationally inefficient. We note here that we do model dispersion in magnitude, but this is done at the level of underlying populations, not observed populations. As shown in \citep{Kessler2013}, the extra colour dispersion shows heavy redshift dependence. As the extra dispersion may arise from underlying supernova physics, rather than Malmquist bias, we decide to incorporate redshift dependence in our extra uncertainty. We thus add $\kappa_0 + \kappa_1 z$ to our observed colour uncertainty (in quadrature), where $\kappa_0$ and $\kappa_1$ are subject to strong Cauchy priors centered on zero and with width $0.002$, to enforce minimum possible extra uncertainty in our model. The $\kappa$ parameters are tightly constrained as they are highly degenerate with the width of the intrinsic colour population $\sigma_c$, and the specific value of this prior was determined from running many iterations of simulation fitting with differing prior strengths on our simple data set with simulated extra dispersion according to the results of \citet{Kessler2013}. The value $0.002$ can be varied by a factor of approximately three in either direction whilst giving stable results, with larger values (more than a factor of three) for the prior resulted in incorrect determination of $\sigma_c$, whilst smaller values resulted in being unable to capture the extra dispersion  Each survey has its own set of $\kappa$ parameters. As such, our combined covariance on the observation $\hat{\eta}$ is given by $\mathcal{C}_{\rm tot} = \mathcal{C} + {\rm DiagMatrix}\left[0, 0, (\kappa_0 + \kappa_1 z)^2\right]$.

Furthermore, we use fiducial simulations (discussed further in Section \ref{sec:selection}) to characterise the shift in the mean colour measured as a function of redshift. {\green This redshift-dependent array is added to our data-set and is included in our modelling by allowing a controlled subtract this shift from our observed data, where the amount of shift subtracted is dictated by parameter $s_c$. Unlike $\kappa_0$ and $\kappa_1$, which are redefined for each survey, $s_c$ represents a global shift in supernovae colour across all surveys. We place a uniform prior on $s_c$.} With the extra dispersion introduced from the $\kappa$ parameters addressing the under-reporting of error, shifting the mean colour address the non-Gaussianity introduced by chromatic smearing. We allow the amount of shift to be parametrised to account for different intrinsic scatter models where parameter $s_c$ controls the fraction of bias subtraction -- 0 for no shift, up to 1 for the maximum shift found in simulations. The scatter models, sourced from \citet{Guy2010} and \citet{Chotard2011} are discussed in detail in Section \ref{sec:simdes}. The maximum shift was found to using the intrinsic scatter model from \citet{Chotard2011}. It is important to note that the effects being modelled and characterised by simulations here are features of the supernova population, not cosmological features, and so the choice of fiducial cosmology has negligible impact.

@@

The underlying supernova population is characterised by underlying distributions in colour, stretch and absolute magnitude. Analytic models often treat the colour and stretch populations independently with a skew normal and normal, respectively. Intrinsic dispersion is treated in a variety of manners in other models, from representing it simply as (Gaussian) scatter on the absolute magnitude of the supernova population, to correlated multivariate normal scatter on the combined magnitude, stretch and colour distribution. Additionally, redshift drift of populations' mean colour and stretch will introduce a cosmological bias in our fits unless the model population possesses similar ability to change as a function of redshift.

@@

On top of the SNIa populations, as described above, we also include a simplistic outlier population that also follows {\rubin} \citep[and therefore ][]{Kunz2007} as a Gaussian mixture; where the mean of the population is fixed to the SNIa population, but the population width is set to a width of $\sigma^{\rm outl} = 1$ in $M_B$, $x_1$ and $c$. With the spectroscopic DES sample, the contamination rate is expected to be far too low to actually fit contamination population, however in future works with photometric samples that will suffer from significantly more contamination, it will be required that extra degrees of freedom are afforded the outlier population. This contamination is not just a result of non-SNIa being observed, but can also arise from host galaxy misidentification and incorrect redshifts - depending on the cuts on the optimised ratio between purity and efficiency, this can result in between 3\% to 9\% misidentification  \citep{Gupta2016}. Proof of concept analysis on simple simulation data show that an acceptable parametrisation to deal with contamination, both from non type Ia supernovae and from catastrophic failure, is to represent the typically dimmer contaminant population as $\langle M_B^{\rm outl} \rangle = \langle M_B \rangle + \delta_{M_B}^{\rm outl}$, where $\delta_{M_B}^{\rm outl}$ is constrained to be positive, or even to be greater than a small positive number to reduce degeneracy between the two populations. For the purposes of the DES spectroscopic sample, which will be confirmed SNIa, $\delta_{M_B}^{\rm outl} =0$. We assume that supernovae fall into the SNIa population as determined by their observed classification probability $\hat{p}$, which is set to unity for the spectroscopic sample.

\subsection{Population Map}

>>>>

\subsubsection{Observables}
Like most of the BHM methods introduced previously, we work from the summary statistics, where each observed supernova has a flux measurement $\hat{m}_B$ (which is analogous to apparent magnitude), stretch $\hat{x}_1$ and colour $\hat{c}$, with uncertainty $\cov$ on those values. Additionally, each supernova has an observed redshift $\hat{z}$ and a host galaxy mass associated with it, $\hat{m}$, where the mass measurement takes the form of a probability of being above $10^{10}$ solar masses. We will also have a probability of each supernovae being a Type Ia, $\hat{p}$. Our set of observables is therefore given as $\lbrace \hat{m}_B, \hat{x}_1, \hat{c}, \hat{z}, \hat{p}, \hat{m}, \cov \rbrace$, as shown in the probabilistic graphical model (PGM) in Figure \ref{fig:pgm}.

@@

\subsubsection{Latent Variables for Observables}

@@

where $\mathcal{C}$ is a covariance matrix reported from light curve fitting.

\subsubsection{Dealing with incorrect observables}

{\green Unfortunately, the assumption of Gaussian uncertainty is incorrect - the errors reported by the light curve fitting process on $m_B$, $x_1$ and $c$ are not accurate. The intrinsic supernovae population is commonly described in terms of coherent magnitude dispersion and in chromatic dispersion, where respectively some supernovae are uniformly brighter or dimmer than the population mean, and the spectral energy density of the supernovae varies non-uniformly as a function of wavelength. The presence of chromatic smearing in the intrinsic supernova population \citep{Guy2010, Chotard2011} means the summary statistics reported from light curve fitting are under-reported, not Gaussian and equation \eqref{eq:pop} is violated. We address this in two ways - increasing the reported uncertainty, and allowing for a global shift in observed colour.}

We model the extra dispersion only in colour, and do so by adding extra independent uncertainty on the colour observation. As shown in \citep{Kessler2013}, the extra colour dispersion shows heavy redshift dependence. As the extra dispersion may arise from underlying supernova physics, rather than Malmquist bias, we decide to incorporate redshift dependence in our extra uncertainty. We thus add $\kappa_0 + \kappa_1 z$ to our observed colour uncertainty (in quadrature), where $\kappa_0$ and $\kappa_1$ are subject to strong Cauchy priors centered on zero and with width $0.002$, to enforce minimum possible extra uncertainty in our model. The $\kappa$ parameters are tightly constrained as they are highly degenerate with the width of the intrinsic colour population $\sigma_c$, and the specific value of this prior was determined from running many iterations of simulation fitting with differing prior strengths on our simple data set with simulated extra dispersion according to the results of \citet{Kessler2013}. The value $0.002$ can be varied by a factor of approximately three in either direction whilst giving stable results, with larger values (more than a factor of three) for the prior resulted in incorrect determination of $\sigma_c$, whilst smaller values resulted in being unable to capture the extra dispersion  Each survey has its own set of $\kappa$ parameters. As such, our combined covariance on the observation $\hat{\eta}$ is given by $\mathcal{C}_{\rm tot} = \mathcal{C} + {\rm DiagMatrix}\left[0, 0, (\kappa_0 + \kappa_1 z)^2\right]$.

Fully covariant extra dispersion on $\lbrace m_B, x_1, c \rbrace$ was also tested, but it showed negligible improvement in recovering unbiased cosmology over just colour dispersion, and was far more computationally inefficient. We note here that we do model dispersion in magnitude, but this is done at the level of underlying populations, not observed populations.

Our second way of addressing the issues with uncertainty are to use fiducial simulations (discussed further in Section \ref{sec:selection}) to characterise the shift in the mean colour measured as a function of redshift. {\green This redshift-dependent array is added to our data-set and is included in our modelling by allowing a controlled subtraction of this shift from our observed data, where the amount of shift subtracted is dictated by parameter $s_c$. Unlike $\kappa_0$ and $\kappa_1$, which are redefined for each survey, $s_c$ represents a global shift in supernovae colour across all surveys. We place a uniform prior on $s_c$.} With the extra dispersion introduced from the $\kappa$ parameters addressing the under-reporting of error, shifting the mean colour address the non-Gaussianity introduced by chromatic smearing. We allow the amount of shift to be parametrised to account for different intrinsic scatter models where parameter $s_c$ controls the fraction of bias subtraction -- 0 for no shift, up to 1 for the maximum shift found in simulations. The scatter models, sourced from \citet{Guy2010} and \citet{Chotard2011} are discussed in detail in Section \ref{sec:simdes}. The maximum shift was found to using the intrinsic scatter model from \citet{Chotard2011}. It is important to note that the effects being modelled and characterised by simulations here are features of the supernova population, not cosmological features, and so the choice of fiducial cosmology has negligible impact.

@@

\subsubsection{Type Ia population} 
The underlying supernova population is characterised by underlying distributions in colour, stretch and absolute magnitude. Analytic models often treat the colour and stretch populations independently with a skew normal and normal, respectively. Intrinsic dispersion is treated in a variety of manners in other models, from representing it simply as (Gaussian) scatter on the absolute magnitude of the supernova population, to correlated multivariate normal scatter on the combined magnitude, stretch and colour distribution. Additionally, redshift drift of populations' mean colour and stretch will introduce a cosmological bias in our fits unless the model population possesses similar ability to change as a function of redshift.

@@

\subsubsection{Outlier populations}

On top of the SNIa populations, as described above, we also include a simplistic outlier population model. We follows {\rubin} \citep[and therefore ][]{Kunz2007} by implementing a Gaussian mixture model. The mean of the outlier population is fixed to the SNIa population, but the population width is set to a width of $\sigma^{\rm outl} = 1$ in $M_B$, $x_1$ and $c$. The probability of each supernovae falling into either population is determined by the observed Type Ia probability $\hat{p}$ discussed in the previous section.

With the spectroscopic DES sample, the contamination rate is expected to be far too low to actually fit contamination population, however in future works with photometric samples that will suffer from significantly more contamination, it will be required that extra degrees of freedom are afforded the outlier population. This contamination is not just a result of non-SNIa being observed, but can also arise from host galaxy misidentification and incorrect redshifts - depending on the cuts on the optimised ratio between purity and efficiency, this can result in between 3\% to 9\% misidentification  \citep{Gupta2016}. Proof of concept analysis on simple simulation data show that an acceptable parametrisation to deal with contamination, both from non type Ia supernovae and from catastrophic failure, is to represent the typically dimmer contaminant population as $\langle M_B^{\rm outl} \rangle = \langle M_B \rangle + \delta_{M_B}^{\rm outl}$, where $\delta_{M_B}^{\rm outl}$ is constrained to be positive, or even to be greater than a small positive number to reduce degeneracy between the two populations. For the purposes of the DES spectroscopic sample, which will be confirmed SNIa, $\delta_{M_B}^{\rm outl} =0$. We assume that supernovae fall into the SNIa population as determined by their observed classification probability $\hat{p}$, which is set to unity for the spectroscopic sample.

\subsection{Mapping function}
